Testing Quasar Unification: I. Radiative Transfer in Clumpy Winds

### 2016-01-28 paper.tex
nearly ready to submit

\maketitle
\begin{abstract} 
Various unification schemes interpret the complex phenomenology of quasars and 
luminous active galactic nuclei (AGN) in terms of a simple picture involving a 
central black hole, an accretion disc and an associated outflow. Here, we continue
our tests of this paradigm by comparing quasar spectra to synthetic spectra of 
biconical disc wind models, produced with our state-of-the-art Monte Carlo radiative 
transfer code. Previously, we have shown that we could produce synthetic spectra 
resembling those of observed broad absorption line (BAL) quasars, but only if the 
X-ray luminosity was limited to 10$^{43}$~erg~s$^{-1}$. 
{\bf Here, we introduce a simple treatment of clumping, and find that a volume filling factor of $\sim0.01$ 
moderates the ionization state sufficently for BAL features to form in the 
rest-frame UV at more realistic X-ray luminosities.}
Our fiducial model shows good agreement with X-ray properties of AGN and the dense, 
X-ray heated wind produces strong line emission in, e.g., \la\ and \civline\ at low inclinations. 
At high inclinations, the synthetic spectra possess prominent LoBAL features. 
Despite these successes, we cannot reproduce all the emission lines seen in quasar 
spectra with the correct equivalent-width ratios, and we find an angular dependence of 
emission-line equivalent width despite the remarkable similarities in the observed emission 
line properties of BAL and non-BAL quasars. Overall, our work suggests that a biconical wind 
can reproduce much of the qualitative behaviour expected from a unified model, but we cannot 
yet provide quantitative matches with quasar properties at all viewing angles. Whether disc 
winds can successfully unify quasars is therefore still an open question.
\end{abstract}

@@

Our kinematic prescription for a biconical disc wind model
follows \cite{SV93}, and is described further by
LK02, H13 and M15. A schematic is shown in Fig.~\ref{fig:cartoon},
with key aspects marked. The general biconical
geometry is similar to that invoked by \cite{MCGV95} and 
\cite{elvis2000} to explain the phenomenonology
of quasars and BALQSOs. {\bf However, we do not include the initial
vertical component present in the Elvis (2000) model, which was
invoked as a qualitative explanation for narrow absorption lines}.

@@

In the SV93 model, a biconical disc wind rises from the accretion 
disc between launch radii $r_{min}$ and $r_{max}$.
The opening angles of the wind are set to $\theta_{min}$ and $\theta_{max}$.
The poloidal velocity along each individual streamline at a poloidal distance $l$ 
is then given by
\begin{equation}
v_l=v_0+\left[v_{\infty}(r_0)-v_0\right]\frac{\left(l/R_v\right)^{\alpha}}{\left(l/R_v\right)^{\alpha}+1},
\label{v_law}
\end{equation}
where $v_0$ is the velocity at the base of the streamline, $\alpha$ is
an exponent governing how quickly the wind accelerations and 
$R_v$ is the `acceleration length', defined as the distance at which
the outflow reaches half of its terminal velocity, $v_{\infty}$.
The terminal velocity is set to a fixed multiple of the escape
velocity, $v_{esc}$, at the base of the streamline (radius $r_0$).
The rotational velocity, $v_{\phi}$, is initially Keplerian ($v_k = [GM/r_0]^{1/2}$),
and the wind conserves specific angular momentum, such that 
\begin{equation}
v_{\phi} r = v_k r_0.
\label{v_law}
\end{equation}
The velocity law is crucial in determining the output spectra,
as it affects not only the projected velocities along the line of sight,
but also the density and ionization state of the outflow.
A wind that accelerates more slowly will have a denser wind base
with correspondingly different ionization and emission characteristics.

@@

The underlying assumption of microclumping is that clump sizes are much smaller than the 
typical photon mean free path, and thus the clumps are 
both geometrically and optically thin. {\bf This approach 
allows one to treat clumps only in terms of their volume filling factor, $f_V$, 
instead of having to specify separately their size and density distributions.
In our model, $f_V$ is independent of position (see section~4.4).
The inter-clump medium is assumed to be unimportant in determining the emergent
spectrum, and is therefore modeled as a vacuum,
although the outflow is still non-porous and axisymmetric.}
The density of the clumps is multiplied by the ``density enhancement'' 
$D=1/f_V$. Opacities, $\kappa$, and emissivities, $\epsilon$, 
can then be expressed as 
\begin{equation}
\kappa = f_V \kappa_C(D);~~\epsilon = f_V \epsilon_C(D).
\end{equation}
Here the subscript $C$ denotes that the quantity is calculated using the 
enhanced density in the clump. The resultant effect is that, {\em for fixed temperature},
processes that are linear in density, such as electron scattering, are unchanged, 
as $f_V$ and $D$ will cancel out. However, any quantity that scales with the square of density, 
such as collisional excitation or recombination, will increase by a factor of $D$.
In our models, the temperature is not fixed, and is instead set by balancing heating and 
cooling in a given cell. In the presence of an X-ray source, this thermal balance is 
generally dominated by bound-free heating and line cooling. The main effect of including 
clumping in our modelling is that it moderates the ionization state due to the increased 
density. This allows an increase in the ionizing luminosity, amplifying the amount of
bound-free heating and also increasing the competing line cooling term
(thermal line emission).

{\bf 
The smallest scale length in a MCRT model such as this is normally the Sobolev length, given by
\begin{equation}
l_S = \frac{v_{th}}{| dv/ds |}
\end{equation}
This is typically $\sim10^{13}$~cm near the disc plane, increasing outwards.
We use the mean density to calculate the Sobolev optical depth, which assumes that
$l_S$ is greater than the typical clump size.
Thus for the microclumping assumption to be formally correct, 
clumps should be no larger than $\sim10^{12}$~cm.
This size scale is not unreasonable for quasar outflows, as
\cite{dekool1995} suggest that BAL flows may have low filling factors with
clump sizes of $\sim10^{11}$~cm. }
% An additional concern with this approach to clumping is that the clumps
% may rapidly cool and expand into the inter-clump medium. This is a well-known
% issue in BLR and outflow models (REFs), and requires some kind of confining mechanism,
% such as magnetic confinement \cite[e.g.][]{dekool1995}. Alternatively, as already discussed,
% the LDI may cause clumpy flows as expected in stellar winds.}

>>>>

\maketitle
\begin{abstract} 
Various unification schemes interpret the complex phenomenology of quasars and luminous active galactic nuclei (AGN) in terms of a simple picture involving a central black hole, an accretion disc and an associated outflow. Here, we continue our tests of this paradigm by comparing quasar spectra to synthetic spectra of biconical disc wind models, produced with our state-of-the-art Monte Carlo radiative transfer code. Previously, we have shown that we could produce synthetic spectra resembling those of observed broad absorption line (BAL) quasars, but only if the X-ray X-ray luminosity was limited to 10$^{43}$~erg~s$^{-1}$. 
 {\bf Here, we introduce a simple treatment of clumping, and find that a filling factor of $\sim0.01$ 
moderates the ionization state sufficently for BAL features to form in the 
rest-frame UV at more realistic X-ray luminosities.} Our fiducial model shows good agreement with AGN X-ray properties and the wind produces strong line emission in, e.g., \la\ and \civline\ at low inclinations. At high inclinations, the spectra possess prominent LoBAL features. Despite these successes, we cannot reproduce all emission lines seen in quasar spectra with the correct equivalent-width ratios, and we find an angular dependence of emission-line equivalent width despite the similarities in the observed emission line properties of BAL and non-BAL quasars. Overall, our work suggests that biconical winds can reproduce much of the qualitative behaviour expected from a unified model, but we cannot yet provide quantitative matches with quasar properties at all viewing angles. Whether disc winds can successfully unify quasars is therefore still an open question.
\end{abstract}

@@

Our kinematic prescription for a biconical disc wind model
follows \cite{SV93}, and is described further by
LK02, H13 and M15. {\bf We emphasise that the purpose of this
purely kinematic wind model is to provide a simple tool for 
exploring different outflow geometries and kinematics, and
the model itself is not hydrodynamic.}
A schematic is shown in Fig.~\ref{fig:cartoon},
with key aspects marked. The general biconical
geometry is similar to that invoked by \cite{MCGV95} and 
\cite{elvis2000} to explain the phenomenonology
of quasars and BALQSOs. {\bf However, we do not include the initial
vertical component present in the Elvis (2000) model, which was
invoked as a qualitative explanation for narrow absorption lines}.

@@

In the SV93 model, a biconical disc wind rises from the accretion 
disc between launch radii $r_{min}$ and $r_{max}$.
The opening angles of the wind are set to $\theta_{min}$ and $\theta_{max}$.
The poloidal velocity along each individual streamline at a poloidal distance $l$ 
is then given by \citep[e.g.][]{rybickihummer1978}
\begin{equation}
v_l=v_0+\left[v_{\infty}(r_0)-v_0\right]\frac{\left(l/R_v\right)^{\alpha}}{\left(l/R_v\right)^{\alpha}+1},
\label{v_law}
\end{equation}
where $v_0$ is the velocity at the base of the streamline, $\alpha$ is
an exponent governing how quickly the wind accelerations and 
$R_v$ is the `acceleration length', defined as the distance at which
the outflow reaches half of its terminal velocity, $v_{\infty}$.
The terminal velocity is set to a fixed multiple of the escape
velocity, $v_{esc}$, at the base of the streamline (radius $r_0$).
The rotational velocity, $v_{\phi}$, is initially Keplerian ($v_k = [GM/r_0]^{1/2}$),
and the wind conserves specific angular momentum, such that 
\begin{equation}
v_{\phi} r = v_k r_0.
\label{v_law}
\end{equation}
The velocity law is crucial in determining the output spectra,
as it affects not only the projected velocities along the line of sight,
but also the density and ionization state of the outflow.
A wind that accelerates more slowly will have a denser wind base
with correspondingly different ionization and emission characteristics.

@@

The underlying assumption of microclumping is that clump sizes are much smaller than the 
typical photon mean free path, and thus the clumps are 
both geometrically and optically thin. {\bf This approach 
allows one to treat clumps only in terms of their volume filling factor, $f_V$, 
instead of having to specify separately their size and density distributions.
In our model, $f_V$ is independent of position (see section~4.4).
The inter-clump medium is modeled as a vacuum,
although the outflow is still non-porous and axisymmetric.
This approach therefore assumes that the inter-clump medium
is unimportant in determing the output spectrum, which
we expect to be true only when density constrasts are large and
the inter-clump medium is both very ionized and of low emissivity and opacity.
}
The density of the clumps is multiplied by the ``density enhancement'' 
$D=1/f_V$. Opacities, $\kappa$, and emissivities, $\epsilon$, 
can then be expressed as 
\begin{equation}
\kappa = f_V \kappa_C(D);~~\epsilon = f_V \epsilon_C(D).
\end{equation}
Here the subscript $C$ denotes that the quantity is calculated using the 
enhanced density in the clump. The resultant effect is that, {\em for fixed temperature},
processes that are linear in density, such as electron scattering, are unchanged, 
as $f_V$ and $D$ will cancel out. However, any quantity that scales with the square of density, 
such as collisional excitation or recombination, will increase by a factor of $D$.
In our models, the temperature is not fixed, and is instead set by balancing heating and 
cooling in a given cell. In the presence of an X-ray source, this thermal balance is 
generally dominated by bound-free heating and line cooling. The main effect of including 
clumping in our modelling is that it moderates the ionization state due to the increased 
density. This allows an increase in the ionizing luminosity, amplifying the amount of
bound-free heating and also increasing the competing line cooling term
(thermal line emission).

{\bf 
The shortest length scale in a Sobolev MCRT treatment such as ours 
is normally the Sobolev length, given by
\begin{equation}
l_S = \frac{v_{th}}{| dv/ds |}
\end{equation}
This is typically $\sim10^{13}$~cm near the disc plane, increasing outwards.
We use the mean density to calculate the Sobolev optical depth, which assumes that
$l_S$ is greater than the typical clump size.
Thus for the microclumping assumption to be formally correct, 
clumps should be no larger than $\sim10^{12}$~cm.
This size scale is not unreasonable for quasar outflows, as
\cite{dekool1995} suggest that BAL flows may have low filling factors with
clump sizes of $\sim10^{11}$~cm. }
% An additional concern with this approach to clumping is that the clumps
% may rapidly cool and expand into the inter-clump medium. This is a well-known
% issue in BLR and outflow models (REFs), and requires some kind of confining mechanism,
% such as magnetic confinement \cite[e.g.][]{dekool1995}. Alternatively, as already discussed,
% the LDI may cause clumpy flows as expected in stellar winds.}
